---
title: "DATA 607 - Project 1 - WNBA ELO Ratings"
author: "Kailot C. Harris"
date: "2026/09/20"
format: html
editor: visual
theme:
  dark: darkly
  light: flatly
---

# Introduction and Overview

In this project, the ELO rating methodology, which was developed to rank Chess players by Arpad Elo, will be applied to WNBA teams. Data will be sourced from the wehoop R package (<https://wehoop.sportsdataverse.org/>). A convention established by the fivethirtyeight blog, and repeated in this article (<https://nicidob.github.io/nba_elo/>) applying ELO ratings to NBA teams, is to start with a baseline value of 1300 for each team.

Anticipated challenges involve maintaining a reasonable scale: WNBA data exists extending back to 1997, and computing 29 seasons of ELO ratings may prove cumbersome if an efficient and repeatable architecture is not implemented from the start. The 2025 season will be used as a baseline from which the methods can be extended backward to cover the entire WNBA history, if practical. Other anticipated challenges may arise when visualizing the results of this analysis. A potential graph could compare ELO ratings vs time for a select group of teams.

# Adaptation

The original project assignment, related to a chess tournament, asked for specific outputs in a .csv file (e.g. player name, player state, total number of points, player pre-rating, and average opponent rating). For this adapted project, the output .csv file will contain columns for: team name, total wins, final ELO rating, and opponents' average ELO rating.

# Codebase

```{r libraries-packages}

#| echo = false
#| warning = false
#| message = false


library(tidyverse)
library(wehoop)
library(elo)
```

## Importing Data from wehoop

To maintain the reduced scope of analyzing only the 2025 season, a request through the `wehoop` R package is made for team-level box-score data in 2025.

```{r data-import}

#| echo = true
#| warning = false
#| message = false

box <- wehoop::load_wnba_team_box(2025)
```

## Tidying and Pruning Data

```{r data-tidying}

#| echo = true
#| warning = false
#| message = false

box <- janitor::clean_names(box)

df <- box |>
  select(game_id, 
         season_type, 
         game_date, 
         game_date_time,
         team_id, 
         team_display_name, 
         team_home_away, 
         team_winner, 
         opponent_team_id, 
         opponent_team_display_name)
```

```{r ELO-init}

#| echo = true
#| warning = false
#| message = false

df <- df |>
    group_by(game_id, team_id) |>
    filter(season_type == 2) |>
    arrange(game_date_time)
```

```{r split-data}

#| echo = true
#| warning = false
#| message = false

away_df <- df |>
  filter(team_home_away == "away") |>
  select(game_id, 
         game_date, 
         away_team_id = team_id, 
         away_team_display_name = team_display_name,
         away_team_winner = team_winner,
         )

home_df <- df |>
  filter(team_home_away == "home") |>
  select(game_id, 
         home_team_id = team_id, 
         home_team_display_name = team_display_name,
         home_team_winner = team_winner
  )
```

```{r join-dfs}

#| echo = true
#| warning = false
#| message = false

full_df <- inner_join(away_df, home_df, by = "game_id") |>
  mutate(home_win = if_else(home_team_winner == TRUE, 1, 0)) |>
  select(game_id,
         game_date,
         home_team_id,
         home_team_display_name,
         away_team_id,
         away_team_display_name,
         home_win
         )

#unique(full_df$home_team_display_name)
# TEAM CLARK present
#unique(full_df$away_team_display_name)
# TEAM COLLIER present

full_df <- full_df |>
  filter(away_team_display_name != "TEAM COLLIER" & home_team_display_name != "TEAM CLARK")

#unique(full_df$home_team_display_name)
# TEAM CLARK removed
#unique(full_df$away_team_display_name)
# TEAM COLLIER removed
```

## Calculating Running ELO Ratings

A running ELO rating is calculated using the ELO R package with documentation here: <https://cran.r-project.org/web/packages/elo/index.html>. The elo.run() function will be used, which is documented in this specific vignette: <https://cran.r-project.org/web/packages/elo/vignettes/running_elos.html>.

```{r ELO-calcs-1}

#| echo = true
#| warning = false
#| message = false

K <- as.integer(50)
default_elo <- as.integer(1300)

elo_run <- elo.run(
  formula = home_win ~ home_team_display_name + away_team_display_name,
  data = full_df,
  k = K,
  initial.elos = default_elo
)
elo_results <- as.data.frame(elo_run)
```

## Tidying ELO Results

```{r elo-tidying}

#| echo = true
#| warning = false
#| message = false

full_elo_df <- bind_cols(full_df, elo_results) 

full_elo_df <- full_elo_df |>
  mutate(
    home_elo_pre =elo.A - update.A,
    away_elo_pre = elo.B - update.B,
    home_elo_post = elo.A,
    away_elo_post = elo.B
  ) |>
  ungroup()

full_elo_df <-full_elo_df |>
  select(game_date, 
         home_team_display_name, 
         away_team_display_name, 
         home_elo_pre, 
         home_elo_post, 
         away_elo_pre, 
         away_elo_post
  )

view(full_elo_df)

```
